---
title: "Causal Forest: Soybean switch to low - cross section"
author: "Sherrie Wang & Jill Deines"
output: 
  html_document:
    toc: true
    toc_float: true
    keep_md: true
---

Goal: Run causal forests on fields which recently switched tillage status; set up as a cross section for 2015 with variable "number of years since switch"

Notes: 

* Many causal forest routines take some time to run (up to a few hours), so this script exports intermediate model results as R Data objects (*.rds) for subsequent use/analysis
* due to randomness inherent in random forest routines, exact results are unlikely to be reproduced

```{r knitrOpts, echo=FALSE}
library(knitr)
opts_chunk$set(cache=FALSE, fig.path='../figure/02.10_causalSoy_toLow/')
```

**R Packages Needed**

```{r packages, message=FALSE, echo=TRUE, eval=TRUE, warning=FALSE}
library(tidyverse)
library(randomForest)
library(earth)
library(corrplot)
library(RColorBrewer)
library(sf)
library(grf)

library(here)

sessionInfo()
```

**Directories**

```{r setDirs}
# output/cleaned data folder
dataFolder <- paste0(here::here(),'/data/tabular_field_data')
dataFileName <- 'soy_switchedFields_toLow_20190616.rds'

# scratch folder for model rdata objects
scratchFolder <- paste0(here::here(),'/data/model_output/causalForest_soy_switchToLow')

# make scratch folder if necessary
dir.create(file.path(scratchFolder), showWarnings = FALSE)
```

# Load data (Jill's code)
Note these variables say 2017 but actual year is 2015 for soy

```{r dataPrep}
crossSection_switchedToLow_2017 <- readRDS(paste0(dataFolder,'/',dataFileName)) 
```


## Propensity scores

```{r define_WXY, eval=FALSE}
# covariates: climate normals and soil variables
X_norm <- crossSection_switchedToLow_2017 %>%
  select(c(contains('norm'), contains('lay1'), 'slope'))

# treatment
W <- crossSection_switchedToLow_2017 %>% dplyr::pull(yearsSinceSwitch)

crossSection_switchedToLow_2017 <- crossSection_switchedToLow_2017 %>%
  rename(W = yearsSinceSwitch,
         Y = yield_tha)

# run
W_forest <- grf::regression_forest(X_norm, W, num.trees = 2000)
W_hat <- predict(W_forest)$predictions

WvarImp <- W_forest %>% 
  variable_importance() %>% 
  as.data.frame() %>% 
  mutate(variable = colnames(W_forest$X.orig)) %>% 
  arrange(desc(V1))

# check propensity scores
p_rf = W_hat
hist(p_rf)

sum(p_rf == 0)
min(p_rf)
sum(p_rf < .005)

sum(p_rf == 1)
max(p_rf)
sum(p_rf > .995)

{plot(smooth.spline(p_rf, W, df = 4))
abline(0, 1)}

# save intermediates
saveRDS(crossSection_switchedToLow_2017, paste0(scratchFolder, '/crossSection_switchedToLow_2017.rds'))
saveRDS(W_forest, paste0(scratchFolder, '/W_forest.rds'))
saveRDS(W_hat, paste0(scratchFolder, '/W_hat.rds'))
saveRDS(WvarImp, paste0(scratchFolder, '/W_forest_varImp.rds'))

```


## Prune Samples
Toss out rows with excessively high/low propensities

### Compare Propensity Filters

```{r propensityFilter, cache = FALSE}
# re-load so chunk runs in markdown output
crossSection_switchedToLow_2017 <- readRDS(paste0(scratchFolder, '/crossSection_switchedToLow_2017.rds'))
W_hat <- readRDS(paste0(scratchFolder, '/W_hat.rds'))
W <- crossSection_switchedToLow_2017 %>% dplyr::pull(W)
p_rf = W_hat



# filter it
overlap0 <- crossSection_switchedToLow_2017 %>%
  mutate(w_hat = W_hat) %>%
  mutate(keep = case_when(w_hat < 0.2 | w_hat > 6 ~ 0,
                        w_hat >= 0.2 | w_hat <= 6  ~ 1))
overlap <- overlap0 %>% filter(keep == 1)
nrow(overlap0) - nrow(overlap)
nrow(overlap)

table(overlap$tillStatus)
table(overlap$year)
table(overlap[,c('tillStatus','year')])

# compare remaining teds
unique(crossSection_switchedToLow_2017$ted)
unique(overlap$ted)


both <- data.frame(W_hat = W_hat,
                   type = paste0('"even" sampling (',
                                 nrow(crossSection_switchedToLow_2017), ' obs)')) %>%
  bind_rows(data.frame(W_hat = overlap$w_hat,
                       type = paste0('Propensity Filtered (', nrow(overlap), ' obs)')))

ggplot(both,
       aes(x = W_hat)) +
  geom_histogram() +
  geom_vline(xintercept = 0.1, col='red') +
  geom_vline(xintercept = 6, col = 'red') +
  facet_wrap(~type, nrow = 2) +
  theme_bw()

{plot(smooth.spline(p_rf, W, df = 4))
abline(0, 1)}

{plot(smooth.spline(overlap$w_hat, overlap$W, df = 4))
abline(0, 1)}

# save overlap data!
saveRDS(overlap, paste0(scratchFolder, '/overlap.rds'))
```


## expected outcome: Y.hat
do this individually too? on overlapped data

### Pick covars

```{r mars_varsel, eval=FALSE}
overlap <- readRDS(paste0(scratchFolder, '/overlap.rds'))

# pick subset of variables by dropping stuff: candidate 1
X_c1 <- overlap %>%
  select(-c(pointID, fips, pxCount, nObs,  W, 
            tillStatus, year_switched, till,
           tillStatus4, keep, w_hat,
            contains('norm'), contains('ted'))) %>%
  tidyr::drop_na()

# mars for var selection
set.seed(5)
marsAll <- earth::earth(Y ~ ., data = X_c1, degree = 2)
summary(marsAll)
ev <- evimp(marsAll) # estimate variable importance
plot(ev)
ev

saveRDS(marsAll, paste0(scratchFolder, '/mars_c1.rds'))
```

### Train Yhat

```{r Yhat, eval = FALSE}
# variables from mars
marsAll <- readRDS(paste0(scratchFolder, '/mars_c1.rds'))
summary(marsAll)
ev <- evimp(marsAll) # estimate variable importance
plot(ev)
ev

envvars <- rownames(ev)


varsToKeep = c(envvars, 'srad_jja')

X <- overlap %>%
  select(varsToKeep)

corr <- cor(X )
corrplot(corr, type = 'upper')

# outcome
Y <- overlap %>% dplyr::pull(Y)

Y_forest <- grf::regression_forest(X, Y, num.trees = 2000)

YvarImp <- Y_forest %>% 
  variable_importance() %>% 
  as.data.frame() %>% 
  mutate(variable = colnames(Y_forest$X.orig)) %>% 
  arrange(desc(V1))
YvarImp


Y_hat <- predict(Y_forest)$predictions

Ychecker <- data.frame(Y = Y,
                       Y_hat = Y_hat,
                       year = overlap$year)
Yagreement <- lm(Y~Y_hat)
summary(Yagreement)

ggplot(Ychecker, aes(x=Y, y = Y_hat)) +
  geom_hex() +
  scale_fill_gradientn(colors = brewer.pal(9,'YlGnBu')) +
  geom_smooth(method= 'lm') +
  coord_equal(xlim=c(0,6), ylim=c(0,6)) + 
  geom_abline(slope = 1, intercept = 0, linetype='dashed') + 
  facet_wrap(~year) +
  theme_bw()

saveRDS(Y_hat, paste0(scratchFolder, '/Y_hat.rds'))
saveRDS(Y_forest, paste0(scratchFolder, '/Y_forest.rds'))
saveRDS(YvarImp, paste0(scratchFolder, '/Y_forest_varImp.rds'))
```


## Run causal forest

```{r cf, eval=FALSE}
# reload overlap
overlap <- readRDS(paste0(scratchFolder, '/overlap.rds'))
Y_hat <- readRDS(paste0(scratchFolder, '/Y_hat.rds'))
WvarImp <- readRDS(paste0(scratchFolder, '/W_forest_varImp.rds'))

# reload mars variables
marsAll <- readRDS(paste0(scratchFolder, '/mars_c1.rds'))
ev <- evimp(marsAll) # estimate variable importance
envvars <- rownames(ev)

# treatment
W <- overlap %>% dplyr::pull(W)
w_hat <- overlap$w_hat

# outcome
Y <- overlap %>% dplyr::pull(Y)

# make covariate list; 
# add in important propensity variables, and things i like (GDD), sherries impt var (tcsoilmapr)
varsToKeep_main = c(envvars,   'tc_soilm_apr',
                   'ppt_apr','awc_lay1_2', 'vpd_july',
                    'meanTemp_apr', 'vpd_july', 'tmax_aug'  )

Xmain <- overlap %>%
  select(varsToKeep_main)

corr <- cor(Xmain )
corrplot(corr, type = 'upper')



cf <- grf::causal_forest(Xmain, Y, W, Y_hat, w_hat, num.trees = 2000)

varImpMain <- cf %>% 
  variable_importance() %>% 
  as.data.frame() %>% 
  mutate(variable = colnames(cf$X.orig)) %>% 
  arrange(desc(V1))


saveRDS(cf, paste0(scratchFolder, '/cf_xc1.rds'))
saveRDS(varImpMain, paste0(scratchFolder, '/cf_xc1_varImp.rds'))

# calculate the partial effect, which takes a small bit of time
ate_cf_aipw <- average_partial_effect(cf)
saveRDS(ate_cf_aipw, paste0(scratchFolder, '/ate_cf_aipw.rds'))
```


# Checking things out

reload rdata stuff

```{r reload}
overlap <- readRDS(paste0(scratchFolder, '/overlap.rds'))

# marsAll <- readRDS(paste0(scratchFolder, '/mars_c1.rds'))


#W_forest <- readRDS(paste0(scratchFolder, '/W_forest.rds'))
#W_hat <- readRDS(paste0(scratchFolder, '/W_hat.rds'))
W_hat <- overlap %>% dplyr::pull(w_hat)
WvarImp <- readRDS(paste0(scratchFolder, '/W_forest_varImp.rds'))

#Y_forest <- readRDS(paste0(scratchFolder, '/Y_forest.rds'))
Y_hat <- readRDS(paste0(scratchFolder, '/Y_hat.rds'))
YvarImp <- readRDS(paste0(scratchFolder, '/Y_forest_varImp.rds'))

cf <- readRDS(paste0(scratchFolder, '/cf_xc1.rds'))
varImp <- readRDS(paste0(scratchFolder, '/cf_xc1_varImp.rds'))
ate_cf_aipw <- readRDS(paste0(scratchFolder, '/ate_cf_aipw.rds'))

# covariates: climate normals and soil variables
X_norm <- overlap %>%
  select(c(contains('norm'), contains('lay1'), 'slope'))

# treatment
W <- overlap %>% dplyr::pull(W)
Y <- overlap %>% dplyr::pull(Y)

constantLess <- readRDS(paste0(scratchFolder, '/crossSection_switchedToLow_2017.rds'))
allWhat <- readRDS(paste0(scratchFolder, '/W_hat.rds'))

# process results ------------------------------
# add predicted treatment effects to data 
constOob_predict <- predict(cf)

# add hats to original data frame
overlapW0 <- overlap %>%
  bind_cols(data.frame(Y_hat = Y_hat))

colNormals <- names(X_norm)
colVars <- varImp %>% dplyr::pull(variable)

overlapW <- overlapW0 %>%
  mutate_at(colNormals, list(Q4n = ~ntile(., 4),
                           Q5n = ~ntile(., 5))) %>%
  mutate_at(colVars, list(Q4 = ~ntile(., 4),
                           Q5 = ~ntile(., 5)))

overlapTau <- overlapW %>%
  bind_cols(constOob_predict) 
```

## propensities

```{r propensity1}
# check propensity scores
p_rf = W_hat
hist(p_rf)

sum(p_rf == 0)
min(p_rf)
sum(p_rf < .005)

sum(p_rf == 1)
max(p_rf)
sum(p_rf > .995)

{plot(smooth.spline(p_rf, W, df = 4))
abline(0, 1)}

WvarImp
```



## Outcome estimation Y hat

```{r yhat_check, fig.width = 5, fig.height = 3, dpi = 600}

YvarImp 

Ychecker <- data.frame(Y = Y,
                       Y_hat = Y_hat,
                       year = overlap$year)
Yagreement <- lm(Y~Y_hat)
summary(Yagreement)

# yield year by year
YcheckerLong <- Ychecker %>%
  tidyr::gather(., key = variable, value = value, Y:Y_hat)


ggplot(YcheckerLong, aes(x=as.factor(year), y = value, group = interaction(year,variable), 
                         fill = variable)) +
  geom_boxplot() +
  theme_bw() +
  ylab('Yield (t/ha)') +
  xlab('Year')

```


# average treatment effects

## variable importance

```{r cf_varimp}
varImp
```


## oob data

```{r ate}
tauhat_rf_aipw = c(ATE=ate_cf_aipw["estimate"],
                   lower_ci=ate_cf_aipw["estimate"] - 1.96 * ate_cf_aipw["std.err"],
                   upper_ci=ate_cf_aipw["estimate"] + 1.96 * ate_cf_aipw["std.err"])
tauhat_rf_aipw


# as a percentage of yield
meanYield <- mean(overlapTau$Y)
meanYield
tauhat_rf_aipw_percent <- tauhat_rf_aipw/meanYield * 100
tauhat_rf_aipw_percent

# sampled in analysis
nrow(overlapTau)
```

## plot treatment effects

```{r cf_predict_oob, fig.height = 3, fig.width = 4, dpi = 300}

ggplot(overlapTau,
       aes(x = predictions)) +
  geom_histogram() +
  xlab('CATE') +
  geom_vline(xintercept = 0, col = 'black', linetype = 'dashed') +
  geom_vline(xintercept = ate_cf_aipw["estimate"], col = 'red') +
  theme_bw()

```
