---
title: 'Fig 1: Study area and treatment propensities'
author: "Jill Deines"
date: "7/12/2019"
output: 
  html_document:
    toc: true
    toc_float: true
    keep_md: true
---

Goal: make Figure 1. Final multi-panel figure assembly done in Inkscape (not provided).

Note the map in Figure 1a is not included but available from the authors of Azzari et al. 2019 (https://www.sciencedirect.com/science/article/pii/S0034425718305157) upon reasonable request.


```{r knitrOpts, echo=FALSE}
library(knitr)
opts_chunk$set(cache=FALSE, fig.path='../figure/03.10_Fig1_studyArea/')
```

**R Packages Needed**

```{r packages, message=FALSE, echo=TRUE, eval=TRUE, warning=FALSE}
library(tidyverse)
library(grf)
library(sf)
library(here)
library(raster)
library(latticeExtra)
library(RColorBrewer)
library(geofacet)

sessionInfo()
```

# Directories

```{r directories}
# scratch folder for model rdata objects: corn causal forest (produced in 01.00)
scratchFolder <- paste0(here::here(),'/data/model_output/causalForest_maize_Longterm')

# gis directory
gisFolder <- paste0(here::here(),'/data/gis')
```

# Reference inset map


```{r stateinset_SOLID, fig.width = 1.75, fig.height = 1.5, dpi = 300, dev = c('png','pdf')}
states <- read_sf(paste0(gisFolder, '/States_continental.shp')) %>%
  st_transform(5070) 

statesTill <- states %>%  dplyr::filter(STATE_ABBR %in%             c('SD','MN','IA','MO','IL','WI','IN','MI','OH'))

till_spdf <- as(statesTill, 'Spatial')
states_spdf <- as(states, 'Spatial')

plot(states[1], col = NA, reset = FALSE, main = NA)
plot(statesTill, col = 'cornflowerblue', add = TRUE)
```


# Figure 1b: conditional probabilities of conservation tillage
treatment propensities on dataset prior to propensity filter - addresses the question "what factors are related to conservation tillage?"

treatment propensities from causal forest subroutine for corn 9 state region

## load causal forest: constant corn

```{r load_cf}
constantLess <- readRDS(paste0(scratchFolder, '/constantLess.rds'))
allWhat <- readRDS(paste0(scratchFolder, '/W_hat.rds'))
WvarImp <- readRDS(paste0(scratchFolder, '/W_forest_varImp.rds'))
```

## format data

```{r propensityProcess}
# add hats to original data frame
prop_all <- constantLess %>%
  bind_cols(data.frame(W_hat = allWhat))

# trim extreme values with few data points to avoid influencing GAM
trimmed <- prop_all %>%
  filter(vpd_jul_norm_hPa > 8 & vpd_jul_norm_hPa < 15,
         temp_apr_norm_C >5,
         awc_lay1_2 >8 & awc_lay1_2 < 24,
         silttotal_lay1_2 < 80,
         claytotal_lay1_2 <= 60,
         sandtotal_lay1_2 <= 80)

# Identify variables to highlight
WvarImp

# select variables and provide better names
varsToPlot <- c('temp_apr_norm_C','pr_may_norm_mm','vpd_jul_norm_hPa', 'awc_lay1_2',
                'slope', 'sandtotal_lay1_2', 'silttotal_lay1_2','claytotal_lay1_2')
betterNames <- c('April Temp','May Precip','July VPD','Soil AWC','Slope', 
                 'Soil: Sand','Soil: Silt','Soil: Clay')
varRenamer <- data.frame(variable = varsToPlot,
                         var2 = betterNames, stringsAsFactors = FALSE) 

trimmed_long <- trimmed %>%
  tidyr::gather(., key = variable, value = value, temp_apr_norm_C:ksat_lay1_2) %>%
  filter(variable %in% varsToPlot) %>%
  left_join(varRenamer)
```

## gam-smoothed plot with data

```{r propensity_4x2_gamHexBins, fig.width = 6.5, fig.height = 4, dpi = 300, dev = c('png','pdf'), bg="transparent"}
# a thing to put soil vars on the bottom row
mygrid2 <- data.frame(
  row = c( 1, 1, 1, 1, 2, 2, 2, 2),
  col = c( 1,2,3,4,1,4,3,2),
  code = c( "April Temp",  "May Precip", "July VPD", "Slope",
            "Soil AWC", "Soil: Clay",  "Soil: Silt",   "Soil: Sand"),
  name = c( "April Temp",  "May Precip", "July VPD", "Slope",
            "Soil: AWC", "Soil: Clay",  "Soil: Silt",   "Soil: Sand"),
  stringsAsFactors = FALSE
)

ggplot(trimmed_long,
       aes(y = W_hat, x = value)) +
  geom_hex(bins = 20) +
  scale_fill_gradientn(colors = brewer.pal(3,'PuBu'),
                       limits = c(1, 4000),
                       na.value = "#2b8cbe") +
  geom_smooth(method = 'gam',
              formula = y ~s(x, bs = 'cs', k=5),
              se = TRUE) +
  facet_geo(~var2, grid = mygrid2, scales = 'free_x') +
  xlab('') + ylab('Probability of Conservation Tillage') +
  ylim(c(0,1)) +
  theme_bw() + theme(axis.text=element_text(size=10),
                     legend.text=element_text(size=10),
                     legend.title = element_blank(),
                     legend.position = 'top',
                     axis.title=element_text(size=11),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     panel.background = element_rect(fill = "transparent",
                                                     colour = NA),
                     plot.background = element_rect(fill = "transparent", 
                                                    colour = NA)) 
```


# Figure 1 A: Tillage Frequency Map
Code for map; will not run (raster not provided)

## Tillage map vis parameters

```{r mapVisParams}
panelBackground <- 'gray30'
stateLines <- 'gray10'
stateFill <- 'gray50'
mapXLim <- c(-450000,1310000)
mapYLim <- c(1570000,2930000)
```


```{r freqMap_scaleRight_9only, eval = FALSE, fig.width = 4.5, fig.height = 3.5, dpi = 600}
percent <-  raster(paste0(gisFolder,'/tillage_percent_full_res500.tif'))
percent[percent < 0] <- NA

percent2 <- mask(percent, till_spdf)

pal <- brewer.pal(9, 'OrRd')
palSmooth <- colorRampPalette(pal)

spplot(percent2, col.regions = palSmooth(20), axes = TRUE,
       colorkey = list(space = 'right'), maxpixels = 1500000,
       xlim = mapXLim, ylim = mapYLim,
       par.settings = list(panel.background = list(col=panelBackground))) +
  latticeExtra::layer(sp.polygons(states_spdf, lwd=.8, col=stateLines)) +
  latticeExtra::layer(sp.polygons(states_spdf, lwd=.8, col=NA, fill = stateFill), under=T)
```




